---
title: "Ps5 Assignment"
author: "Tawsif Ahmed"
date: "`r Sys.Date()`"
output: html_document
---

# Ps5: Rmarkdown, basic plots

### Load and check data

1.
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

2. 
```{r, message=FALSE}
gap <- read_delim("Data/gapminder.csv.bz2")
gap %>% 
  dim()
```
There is `r nrow(gap)` rows and `r ncol(gap)` columns in the data.

3. 

```{r}
gap %>% 
  sample_n(10)
```
There are no issues with the data. 

### Descriptive Statistics 

1. 

```{r}
gap %>% 
  filter(!is.na(name) | !is.na(iso2) | !is.na(iso3)) %>% 
  summarize(n1 = n_distinct(name), n2 = n_distinct(iso2), 
            n3 = n_distinct(iso3))
```
2. a)
```{r}
gap %>% 
  group_by(iso2) %>% 
  filter(!is.na(iso2)) %>% 
  summarise(n = n_distinct(name)) %>% 
  arrange(desc(n))
```
For each iso-2 code, there are only one name and doesn't correspond to more than one.

b)
```{r}
gap %>% 
  group_by(name) %>% 
  summarise(n = n_distinct(iso3)) %>% 
  arrange(desc(n))
```
For iso3 code, there is **4** iso3 codes that does not correspond to any countries. Thus, it has more iso3 codes than names.  

3. 
```{r}
gap %>% 
  filter(!is.na(time)) %>% 
  summarize(min = min(time),max = max(time))
```
The min year is **1960** and the max year is **2019**.

### CO2 emissions 

1. 
```{r}
gap %>% 
  group_by(time) %>% 
  summarize( n = sum(is.na(co2)) + sum(is.na(co2_PC)) ) %>% 
  arrange(desc(n))
```
*2017-2019* all of these years have equal amount of most missing values for CO2 emissions.

2. 
```{r}
gap %>% 
  filter(name %in% c("United States of America","China",
                     "India", "France", "Canada")) %>% 
  filter(!is.na(co2)) %>% 
  group_by(name) %>% 
  ggplot(aes(x = time, y = co2, col = name))+
  geom_line() + ggtitle("CO2 emissions over time")+
  xlab("Year") + ylab("CO2 emissions (kt)") +
  theme(plot.title = element_text(hjust = 0.5))
```


The total amount of CO2 emissions roughly increases for all countries from 1960 to 1980. Some countries keep their
CO2 emissions the same to the present time expect **China** and **India**. Both's total emissions have increased while China's increased significantly compared to the rest.  

3.
```{r}
gap %>% 
  filter(name %in% c("United States of America","China",
                     "India", "France", "Canada")) %>% 
  filter(!is.na(co2_PC)) %>% 
  group_by(name) %>% 
  ggplot(aes(x = time, y = co2_PC, col = name))+
  geom_line() + ggtitle("CO2 per capita emissions over time")+
  xlab("Year") + ylab("CO2 emissions (mt per capita)") +
  theme(plot.title = element_text(hjust = 0.5))
```


According to the figure, **United States** had the most CO2 emissions per capita compared to the rest. All countries are decreasing in over time expect *China* and *India*, which are increasing. 

4. 
```{r}
gap %>% 
  group_by(region) %>% 
  filter(time == "2016" & !is.na(region)) %>% 
  summarize(CO2 = mean(co2_PC, na.rm= TRUE)) %>% 
  arrange(desc(CO2))
```
Since *2017-2019* had most missing values regarding co2 emissions, I used 2016 as the recent year. So in 2016, **Europe** had most CO2 emission per capita while **Africa** had the least CO2 emission per capita.

5.
```{r}
gap %>% 
  group_by(region,time) %>% 
  filter(time %in% c("1960","2016") & !is.na(region)) %>% 
  summarize(CO2 = mean(co2_PC, na.rm= TRUE)) %>% 
  ggplot(aes(x = region , y = CO2, fill = factor(time))) +
  geom_col(position = "dodge") +
  labs(x = "Continent", y ="Average CO2 per capita") +
  ggtitle("CO2 per capita 1960 vs. 2016") +
  theme(plot.title = element_text(hjust = 0.5))
  
```


Out of all of the continents, **Americas** had a decrease in CO2 per capita from *1960* to *2019* while other ones had an increase. **Asia** had the most increase in CO2 per capita from *1960* to *2019*. 

6. 
```{r}
gap %>% 
  group_by(region, name) %>% 
  filter(time == 2016) %>%
  filter(!is.na(region) & !is.na(co2_PC) & !is.na(name)) %>% 
  summarize(n = co2_PC) %>% 
  filter(rank(desc(n)) <= 3 | rank(n) <= 3) %>% 
  arrange(desc(n)) %>% 
  arrange(region)
```

### GDP per capita

1. 
```{r}
gap %>% 
  filter(!is.na(lifeExpectancy) & !is.na(GDP_PC)) %>% 
  filter(time == 1960) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, col = region, size = totalPopulation)) +
  geom_point() + labs(x = "GDP per capita", y ="Life Expectancy ") +
  ggtitle("Life Expectancy vs. GDP per capita in 1960") +
  theme(plot.title = element_text(hjust = 0.5))
```


Based on the graph, there are a large amount of countries that seem to have life-expectancies all over the place when the GDP per capita is below *$10000*. After *$10000* in GDP per capita, the relationship becomes the larger GDP per capita, the higher the life-expectancies. 

2. 
```{r}
gap %>% 
  filter(!is.na(lifeExpectancy) & !is.na(GDP_PC)) %>% 
  filter(time == 2019) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, col = region, size = totalPopulation)) +
  geom_point() + labs(x = "GDP per capita", y ="Life Expectancy ") +
  ggtitle("Life Expectancy vs. GDP per capita in 2019") +
  theme(plot.title = element_text(hjust = 0.5))
```


3. Based on the graph, the life expectancies has increased overall from 1960 regardless of GDP per capita. The relationship is still the same after *$10000* in GDP per capita, where life expectancies increases with GDP per capita. However, the life expectancies does not go higher than **90 years** in any countries. 

4. 
```{r}
gap %>% 
  filter(time %in% c(1960,2019) & !is.na(lifeExpectancy)
         & !is.na(region)) %>% 
  group_by(region,time) %>% 
  summarize(avg_LE = mean(lifeExpectancy))
```
The results fit with the figures since it shows that the life expectancy has increased overall regardless of region, which is true for the result as well. 

5. 

```{r}
gap %>% 
  filter(!is.na(lifeExpectancy) & !is.na(region)) %>% 
  group_by(region, time) %>% 
  summarize(avg_LE = mean(lifeExpectancy)) %>% 
  mutate(growth = avg_LE - lag(avg_LE)) %>% 
  arrange(desc(growth))
```
Out of all continents, **Asia** and **Africa** had the highest growth in life expectancy couple of times. **Africa** had most recent highest growth which could be an effect from foreign aid and advance in  technologies. 

6. 
```{r}
gap %>% 
  filter(time %in% c(1960,2019) & !is.na(GDP_PC)) %>%
  ggplot(aes(x = GDP_PC , fill = factor(time))) +
  geom_histogram( position = "dodge") +
  labs(x = "GDP per capita", y ="Number of Countries") +
  ggtitle("GDP per capita in 1960 vs. 2019") +
  theme(plot.title = element_text(hjust = 0.5)) 

```


In 1960, majority of the countries were below *$30000* GDP per capita. While in 2019, there are many countries that suppressed the *$30000* GDP per capita. Some going higher than *$90000* GDP per capita. For 2019, there are more countries' GDP available to be accounted from 1960.

7.

```{r}
gap %>% 
  group_by(time) %>% 
  filter(!is.na(lifeExpectancy) & !is.na(name),time %in% c(1960,2019))%>%
  mutate(pos = rank(desc(lifeExpectancy))) %>% 
  select(name, lifeExpectancy, time,  pos) %>% 
  filter(name == "United States of America")
```
United States ranked lower in life expectancy even though the life expectancy got higher. 

8. 
```{r}
gap %>% 
  group_by(time) %>% 
  filter(!is.na(lifeExpectancy) & !is.na(name),time %in% c(1960,2019))%>%
  mutate(pos = rank(desc(lifeExpectancy))) %>% 
  select(name, lifeExpectancy, time,  pos) %>% 
  mutate(relative_rank = pos/sum(n())) %>% 
  filter(name == "United States of America") 
```
Based on the relative rank, United States did still fall down in the ranking, but not significantly. 

### Finally

I spend around 6-7 hours on this PS. 